---
title: "FinalProjectCode"
format: html
---
# Load in Necessary Packages

```{r}
## Load Necessary Packages
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(reshape2)
library(ggplot2)
```

# Load In First Dataset

```{r}
# Step 1: Read CSV without column names
mlb_data_raw <- read_csv("https://raw.githubusercontent.com/nwt5144/FinalProjectStat184/refs/heads/main/stats.csv", show_col_types = FALSE)
```


# Fix All Column Names In Original Dataset and Calculate Summary Statistics 

```{r}
# Create Summary Statistics Table ----
batting_data <- mlb_data_raw %>%
  select(
    'last_name, first_name',
    home_run, 
    walk, 
    k_percent, 
    b_intent_walk, 
    avg_swing_speed, 
    launch_angle_avg,
    hard_hit_percent,
    whiff_percent
  )



# Calculate summary statistics
summary_table <- batting_data %>%
  reframe(
    Variable = c(
      "home_run", 
      "walk", 
      "k_percent", 
      "b_intent_walk", 
      "avg_swing_speed", 
      "launch_angle_avg",
      "hard_hit_percent",
      "whiff_percent"
    ),
    Mean = c(
      mean(home_run, na.rm = TRUE), 
      mean(walk, na.rm = TRUE), 
      mean(k_percent, na.rm = TRUE), 
      mean(b_intent_walk, na.rm = TRUE), 
      mean(avg_swing_speed, na.rm = TRUE), 
      mean(launch_angle_avg, na.rm = TRUE),
      mean(hard_hit_percent, na.rm = TRUE), 
      mean(whiff_percent, na.rm = TRUE)
    ),
    Median = c(
      median(home_run, na.rm = TRUE), 
      median(walk, na.rm = TRUE), 
      median(k_percent, na.rm = TRUE), 
      median(b_intent_walk, na.rm = TRUE), 
      median(avg_swing_speed, na.rm = TRUE), 
      median(launch_angle_avg, na.rm = TRUE),
      median(hard_hit_percent, na.rm = TRUE), 
      median(whiff_percent, na.rm = TRUE)
    ),
    Minimum = c(
      min(home_run, na.rm = TRUE), 
      min(walk, na.rm = TRUE), 
      min(k_percent, na.rm = TRUE), 
      min(b_intent_walk, na.rm = TRUE), 
      min(avg_swing_speed, na.rm = TRUE), 
      min(launch_angle_avg, na.rm = TRUE),
      min(hard_hit_percent, na.rm = TRUE), 
      min(whiff_percent, na.rm = TRUE)
    ),
    Maximum = c(
      max(home_run, na.rm = TRUE), 
      max(walk, na.rm = TRUE), 
      max(k_percent, na.rm = TRUE), 
      max(b_intent_walk, na.rm = TRUE), 
      max(avg_swing_speed, na.rm = TRUE), 
      max(launch_angle_avg, na.rm = TRUE),
      max(hard_hit_percent, na.rm = TRUE), 
      max(whiff_percent, na.rm = TRUE)
    )
  )
```


# Find and Show Summary Statistics for First Dataset

```{r}
# Display Summary Statistics Table ----
summary_table %>%
  kable(
    caption = "Summary Statistics for Key Batting Variables", 
    digits = 2, 
    align = c("l", rep("c", 4))
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE, 
    font_size = 14
  ) %>%
  row_spec(0, bold = TRUE)  # Bold header row
```


# Scatterplot: Home Run vs Average Swing Speed

```{r}
# Scatterplot: Home Run vs Average Swing Speed
ggplot(batting_data, aes(x = avg_swing_speed, y = home_run)) +
  geom_point(color = "blue", alpha = 0.7) +
  labs(
    title = "Home Runs vs Average Swing Speed",
    x = "Average Swing Speed",
    y = "Home Runs"
  ) +
  theme_minimal()
ggsave("plots/swing_vs_hr.png", width = 7, height = 5)

```


# Scatterplot: K% vs Hard Hit %

```{r}
# Scatterplot: K% vs Hard Hit %
ggplot(batting_data, aes(x = k_percent, y = hard_hit_percent)) +
  geom_point(color = "red", alpha = 0.7) +
  labs(
    title = "Strikeout % vs Hard Hit %",
    x = "Strikeout %",
    y = "Hard Hit %"
  ) +
  theme_minimal()
```


# Scatterplot: K% vs Hard Hit %

```{r}
# Scatterplot: K% vs Hard Hit %
ggplot(batting_data, aes(x = k_percent, y = home_run)) +
  geom_point(color = "red", alpha = 0.7) +
  labs(
    title = "Strikeout % vs Home Runs",
    x = "Strikeout %",
    y = "Home Runs"
  ) +
  theme_minimal()
```


# Scatterplot: K% vs Hard Hit %


```{r}
# Scatterplot: K% vs Hard Hit %
ggplot(batting_data, aes(x = avg_swing_speed, y = whiff_percent)) +
  geom_point(color = "red", alpha = 0.7) +
  labs(
    title = "Average Swing Speed vs Whiff %",
    x = "Average Swing Speed",
    y = "Whiff %"
  ) +
  theme_minimal()
```


# Correlation matrix after removing Player name columns

```{r}
# Compute correlation matrix after removing Player name columns
correlation_matrix <- batting_data %>%
  select(-`last_name, first_name`) %>%  # drop non-numeric columns
  cor(use = "complete.obs")

# Melt correlation matrix into long format
correlation_melt <- melt(correlation_matrix)

# Plot heatmap
ggplot(correlation_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", 
    high = "red", 
    mid = "white", 
    midpoint = 0, 
    limit = c(-1, 1), 
    space = "Lab", 
    name = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(
    title = "Correlation Heatmap",
    x = "",
    y = ""
  )
```

# Prepare batting_data for merging with second dataset

```{r}
# Prepare batting_data for merging
batting_data_for_combo <- batting_data %>%
  separate(`last_name, first_name`, into = c("LastName", "FirstName"), sep = ", ", remove = FALSE) %>%
  select(-`last_name, first_name`) %>%
  mutate(
    Player = paste(FirstName, LastName),
    Player = str_trim(Player)
  )

# Read roster data
roster_data <- read_csv("https://raw.githubusercontent.com/nwt5144/FinalProjectStat184/refs/heads/main/NlAlPos.csv")

# Clean roster_data
roster_data <- roster_data %>%
  mutate(
    Player = str_replace(Player, "\\*", ""), 
    Player = str_trim(Player)
  )

# Keep only necessary columns
roster_data_small <- roster_data %>%
  select(Player, Lg, Pos)

# Merge on Player
combined_data <- batting_data_for_combo %>%
  left_join(roster_data_small, by = "Player")

# Clean and group Pos
combined_data <- combined_data %>%
  mutate(
    Pos = str_replace_all(Pos, "\\*", ""),
    Pos = str_extract(Pos, "\\d"),
    Pos = case_when(
      Pos == "1" ~ "Pitcher",
      Pos %in% c("2", "3", "4", "5", "6") ~ "Infielder",
      Pos %in% c("7", "8", "9") ~ "Outfielder",
      TRUE ~ NA_character_
    )
  )

head(combined_data)
```


# Scatterplot: Home Runs vs Walks by League


```{r}
# Scatterplot: Home Runs vs Walks by League
ggplot(combined_data, aes(x = walk, y = home_run, color = Lg)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(
    title = "Home Runs vs Walks by League",
    x = "Walks",
    y = "Home Runs"
  ) +
  theme_minimal()
```


# Scatterplot: Home Runs vs Hard Hit % by Position Group

```{r}
# Scatterplot: Home Runs vs Hard Hit % by Position Group
ggplot(combined_data, aes(x = hard_hit_percent, y = home_run, color = Pos)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(
    title = "Home Runs vs Hard Hit % by Position",
    x = "Hard Hit %",
    y = "Home Runs"
  ) +
  theme_minimal()
```


# Scatterplot: Home Runs vs Walks by League and Position

```{r}
# Scatterplot: Home Runs vs Walks by League and Position
ggplot(combined_data, aes(x = walk, y = home_run, color = Pos)) +
  geom_point(alpha = 0.7, size = 2) +
  facet_wrap(~ Lg) +
  labs(
    title = "Home Runs vs Walks by League and Position",
    x = "Walks",
    y = "Home Runs"
  ) +
  theme_minimal()
```

